\documentclass{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bibref}
\usepackage{authblk}
\usepackage{hyperref}
\title{WORDBOX: A case study in SAT Encodings}
\author{Victor S. Miller}
\affil{SRI, Computer Science Laboratory, Menlo Park, CA 94025
\footnote{Much of the this work performed at the IDA Center for Communications Research, Princeton, NJ 08540
USA}}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\begin{document}
\maketitle
\begin{abstract}
  The great effectiveness of SAT solvers for large, difficult, SAT
  problems is a recent welcome development.  It is well know to
  practitioners in that field that the encoding of a problem can have
  a significant and radical effect on the practical solution time.

  In particular, many combinatorial problems have a straightforward
  ``easy'' encoding which is logically sufficient to specify the
  problem.  However, SAT solvers are very poor at inferring ``global''
  knowledge about the problem.  In fact, with some problems, such as
  the ``Pigeon Hole Principal'', in which global knowledge makes the
  problem trivial, branching SAT solvers must take exponential time.
  Examples of global knowledge are graph connectedness and
  parity. Therefore, in order to find an improved encoding global
  knowledge should be used.

  In this note we consider the encoding a recreational Mathematics
  problem known as \emph{Wordbox}.  One is given a list of words, and
  an $m \times n$ rectangular grid.  The object of the problem is to
  label the grid points with letters so that one can trace out all of
  the words in the list by moving adjacent grid points (using up/down,
  right/left moves).  We show that this problem is an instance of the
  \emph{labeled graph homomorphism} problem, in which we're given two
  finite undirected graphs $G$ and $H$, along with labels $\ell(v)$ for
  every node of $G$.  The object of the problem is to find a map
  $f: V(G) \rightarrow V(H)$ such that if $(v,v') \in E(G)$ then
  $(f(v), f(v')) \in E(H)$ (this is a \emph{graph homomorphism}), and
  a labeling $\ell(w)$ of $w \in V(H)$ such that, for all $v \in V(G)$
  we have $\ell(f(v)) = \ell(v)$.  In turn, we show that the labeled
  graph homomorphism problem is a subset of the \emph{colored graph
    homomorhphism problem}.

  We investigate a number of different encodings of the global
  knowledge and their effect on solving times for various SAT solvers.
\end{abstract}

\section{A recreational problem}
\label{sec:recreational}

In the GCHQ \emph{Kryptos Kristmas Kwiz} of 2006
\url{https://theintercept.com/2015/12/25/gchq-play-a-british-spy-game/}
\footnote{Solutions at \url{https://theintercept.com/2016/01/01/gchq-play-a-british-spy-game-the-solutions/}}
describes a problem called ``wordbox'':
\begin{quotation}
  A wordbox is a rectangle of letters in which a series of words can
  be read by starting at one letter and moving to an adjacent letter
  in a row or column, but not diagonally. Retracing steps and reusing
  letters is allowed.  Thus a wordbox might be
  \begin{center}
    \begin{tabular}[h]{|c|c|c|c|}
      \hline
       D & O & R & F \\
      \hline
       I & G & E & S \\
      \hline
       T & A & O & T \\
      \hline
    \end{tabular}
  \end{center}
  which contains {\tt DODO, DOG, FROG, STOAT TIGER} and {\tt TIT}, but
doesn't contain {\tt GOAT} or {\tt GEESE}. This wordbox contains 12
letter cells, so its area is 12.  What is the smallest (in area)
wordbox that includes the planets: {\tt MERCURY, VENUS, EARTH, MARS,
  JUPITER, URANUS, NEPTUNE, PLUTO}?
\end{quotation}
The answer given (without any proof of minimality) is:
\begin{center}
  \begin{tabular}[h]{|c|c|c|c|}
    \hline
    V & E & A & M \\
    \hline
    A & N & R & T \\
    \hline
    R & U & S & H \\
    \hline
    N & T & A & \quad \\
    \hline
    E & P & I & M \\
    \hline
    L & U & T & E \\
    \hline
    P & J & O & R \\
    \hline
    Y & R & U & C \\
    \hline
  \end{tabular}
\end{center}
\section{A logical statement}
\label{sec:logical-statement}

To solve the original recreational problem, we pose it as a problem of
\emph{Labeled graph homomomorphism}:

\begin{definition}[Graph Homomorphism]
  Let $G, H$ be graphs.  A \emph{graph homomorphism} from $G$ to $H$
  is a map $\phi : V(G) \rightarrow V(H)$ such that, for all $(v,w) \in
  E(G)$, we have $(\phi(v), \phi(w)) \in E(H)$.
\end{definition}
\begin{definition}[Labeled Graph]
  Let $G$ be a graph.  A \emph{labeling} of $G$ is a map $\ell: V(G)
  \rightarrow C$, for some set $C$.
\end{definition}
\begin{definition}[Labeled homomorphism]
  Let $G$ be a labeled graph, with a \emph{labeling} $\ell: V(G)
  \rightarrow C$, and $H$
  be a graph.  A \emph{labeled homomorphism} is a homomorphism $\phi:
  V(G) \rightarrow V(H)$, and a labeling $\ell': V(H) \rightarrow C$,
  such that $\ell'(\phi(v)) = \ell(v)$ for all $v \in V(G)$.
\end{definition}

For the original problem, the graph $G$ is a disjoint union of simple
paths: for each word, of length $k$, there is a path $v_1, \dots,
v_k$, where the edges are $(v_i, v_{i+1})$ for $i=1, \dots, k-1$, and
$\ell(v_i)$ is the $i$-th letter in the word.  The graph $H$ is an $m
\times n$ \emph{grid graph}: its vertices are pairs $(i,j), 0 \le i <
m, 0 < j < n$, with $((i,j), (i',j'))$ an edge if and only if
$|i-i'| + |j-j'| = 1$.

We may state the problem as follows: We are given two
undirected graphs $G, H$, and a \emph{labeling} $\ell : V(G)
\rightarrow C$.  We want to know if there exists a graph homomorphism
$f: V(G) \rightarrow V(H)$, and a labeling $\ell': V(H) \rightarrow C$
such that $\ell'(f(v)) = \ell(v)$ for all $v \in V(G)$.


For the original problem, the graph $G$ is a disjoint union of paths,
$H$ is the \emph{grid graph}
\begin{equation}
  \label{eq:1}
  \begin{aligned}
    \exists & f: V(G) \rightarrow V(H), \ell': V(H) \rightarrow C \\
    \forall & (v,v') \in E(G), (f(v), f(v')) \in E(H) \\
    \wedge \forall & v \in V(G), \ell'(f(v)) = \ell(v).
  \end{aligned}
\end{equation}
where $f$ is a function.  However, in order to encode this in
quantifier normal form we need to encode $f$ as a binary relation.

So $w=f(x)$ means
\begin{displaymath}
  \forall x \exists w R(x,w) \wedge \forall x,w,w' ((w = w') \vee
  \neg R(w,x) \vee \neg R(x,w')).
\end{displaymath}

\section{The colored homomorphism problem}
\label{sec:colored}

\begin{definition}[Edge Colored Graph]
  An \emph{edge colored graph} is a finite undirected graph $G =
  (V(G), E(G))$ a set, $C$ of colors and a map $\phi: E(G) \rightarrow
  C$, which assigns a color to each edge.
\end{definition}
\begin{definition}[Edge Colored Homomorphism]
Let $G$ and $H$ be two edge colored graphs with the same set of
colors.  An \emph{edge colored homomorphism} is a homomorphism $f : G
\rightarrow H$, such that, for all edges $e \in E(G)$, the color of
the edge $f(e) \in E(H)$ is the same as that of $e$.
\end{definition}

We now reduce the vertex colored homomorphism problem to the 2-colored
edge homomorphism problem.  Given a vertex colored graph $G$,
construct the edge colored graph $G'$ with two colors, red and green.
The edges of $G'$ will be a copy of the edges of $G$ colored green.
For each $v \ne v' \in V(G)$ such that $\ell(v) \ne \ell(v')$, and
$(v,v') \not \in E(G)$, we add an edge $(v,v')$ colored red.  For the
target graph $H$, we construct a 2-colored graph $H'$, whose green
edges are a copy of the edges of $H$, and whose red edges are
$(v,v') \not \in H$, for $v \ne v'$.
\begin{proposition}
  There is a vertex colored homomorphism from $G$ to $H$ if and only
  if there is an edge colored homomorphism from $G'$ to $H'$.
\end{proposition}

\section{Automorphisms of edge colored graphs}
\label{sec:automorphisms}

Programs such as {\tt nauty} can calculate the automorphism group of a
vertex colored graph.  That is, if $G$ is a vertex colored graph, with
color function $c$, and \emph{automorphism} $\phi: G \Rightarrow G$ is
a one-to-one map $\phi: V(G) \rightarrow V(G)$ such that, for each
edge $\{v_1, v_2\} \in E(G)$, we have $\{\phi(v_1), \phi(v_2)\} \in
E(G)$ and for each $v \in V(G)$ we have $c(v) = c(\phi(v))$.

Given an edge two colored graph, $G$, we form a new vertex colored
graph $G'$ whose vertices are the disjoint union of two copies of
$V(G)$: $V(G) \times \{0,1\}$.  The edges of $G'$ are as follows:
for each $v \in V(G)$ there is an edge $\{(v,0), (v,1)\}$.  If an edge
$\{v,w\} \in E(G)$ has color $c$, there is an edge $\{(v,c), (w,c)\}
\in E(G')$.  Vertices of the form $(v,c)$ have color $c$.  If $\phi$
is an automorphism of $G'$, then necessarily $\phi((v,c)) = (w,c)$ for
some $w$, by preserving colors.  If $\phi((v,0)) = (w,0)$ then we must
have $\phi((v,1)) = (w,1)$ for the same $w$ because the only edges
going from $(G,0)$ to $(G,1)$ are the vertical edges.

\section{Fractional Homomorphism}
\label{sec:fractional}

Let $G$ and $H$ be undirected graphs.  The \emph{incidence matrix} for
$G$ (resp. $H$) is a matrix whose rows are indexed by $V(G)$ and
columns by $E(G)$, with $M_{v,e}^G = [v \in e]$.  We define two
matrices $X$ and $Y$ which will encode a homomorphism.  Given $v \in
V(G), w \in V(H)$, define $X_{w,v} = 1$ if $v$ maps to $w$ and 0
otherwise.  Similarly, given $e \in E(G)$ and $f \in E(H)$ define
$Y_{f,e} = 1$ if $e$ maps to $f$ and 0 otherwise.
\begin{proposition}
  Given $G, H$ simple undirected graphs, there is a homomorphism from
  $G$ to $H$, if and only if there are 0/1 matrices $X_{w,v}$ for
  $v \in V(G), w \in V(H)$, and $Y_{f,e}$ for $e \in E(G), f \in E(H)$
  with $\sum_{w \in V(H)} X_{w,v} = 1$ and
  $\sum_{f \in E(H)} Y_{f,e} = 1$ and $X M^G = M^H Y$.
\end{proposition}
\begin{proof}
  Note that
\end{proof}
\begin{equation}
\label{eq:vertex:map}
(X M^G)_{w,e} = \sum_{v \in V(G)} X_{w,v} M^G_{v,e} = \sum_{v \in e}
X_{w,v},
\end{equation}
and
\begin{equation}
\label{eq:edge:map}
  (M^H Y)_{w, e} = \sum_{f \in E(H)} M^H_{w,f} Y_{f,e} = \sum_{f \in
    E(H), w \in f} Y_{f,e}.
\end{equation}
We can interpret equation~\eqref{eq:vertex:map} as saying that $(X
M^G)_{w,e}$ is the number of endpoints of edge $e$ that map to
$w$, which is 0, 1 or 2.  We can also interpret
equation~\eqref{eq:edge:map} as saying that $(M^H Y)_{w,e}$ is
the number of endpoints of $e$ that map to $w$.  However, this
quantity is 0 or 1, since the graphs are simple.  Thus, if the two
sides are equal there is a homomorphism.  The converse is clear.

We can express homomorphism in terms of these matrices:
$X M^G = M^H Y$.  Note that
where $w \in V(H)$, and

The first equation encodes mapping on vertices $v \rightarrow w$, and
the second encodes mapping on edges $e \rightarrow f$.  If we require
that the $X$ and $Y$ are left stochastic (row sum = 1, nonnegative
entries), this is a relaxation.

More specifically, we would like $X$ to have the property that
$X_{w,v} = 1$ if $v$ maps to $w$ and 0 otherwise.  In addition, we
would like $Y_{f,e} = 1$ if $e$ maps to $f$ and 0 otherwise.

If this is the case then $(X M^G)_{w,e}$ is the number of vertices of
$G$ incident to $e$ that map to $w$, which is the number of endpoints
of $e$ that map to $w$.  Notice that for a simple graph this is 0, 1
or 2.  However, as we shall see in the next paragraph.  The entry on
the right hand side must be 0 or 1.

We also have $(M^H Y)_{w,e}$ is 1 if $e$ maps to an edge incident to
$w$ and 0 otherwise.  Thus, the two sides are equal if and only if
there is a homomorphism from $G$ to $H$.  If the edges of $G$ and $H$
are colored, and we require that colors map to like colors, we can
enforce this by restricting elements of the matrix $Y$.

We have an LP (linear programming) relaxation by requiring that the
matrices be left stochastic.  That is $1 \ge Y_{f,e} \ge 0$ and
$\sum_{e \in E(G)} Y_{f,e} = 1$.  We can then interpret the value
$Y_{f,e}$ as the probability that the edge $e$ is mapped to the edge
$f$.  Similarly, we require that $X$ is left stochastic.  In this case
we can interpret $X_{w,v}$ as the probability that $v$ maps to $w$.
\end{document}
